---
title: "SSPs Human Capitals - National"
author: "James Millington"
format: pdf
editor: visual
---

## Libraries

```{r output=FALSE}
library(tidyr)
library(dplyr)
library(readxl)
library(ggplot2)
library(ggrepel)
library(ggiraph)
library(viridisLite)
```

## Load Data

```{r}
load_ssp <- function(filepath,varlab){
  #for each SSP data sheet
  for(n in 1:5){
    ssp <- paste0("ssp", n)
    sheet <- read_excel(filepath, sheet = ssp, na='<Null>')
    
    #pivot longer, add ssp id column
    lsheet <- sheet %>%
      pivot_longer(
        cols = !OID:unit,
        names_to='year',
        values_to='value'
      ) %>%
      mutate(ssp=n)
    
    #calculate ranks of values 
    lranks <- lsheet %>%
      group_by(year) %>%
      mutate(unit='rank',
             #value=min_rank(desc(value)))   #high rank number is low value
             value=min_rank(value))   #high rank number is high value
  
    #combine into single dataframe  
    ldat <- rbind(lsheet,lranks)
    if(n > 1){
      alldat <- rbind(alldat,ldat)
    } else { alldat <- ldat}
  }
  alldat <- mutate(alldat,variable=varlab)
  return(alldat)
}
```

```{r}
edu <- load_ssp("data/edu.xlsx", "edu")
ma <- load_ssp("data/ma.xlsx", "ma")
gdp <- load_ssp("data/gdp.xlsx", "gdp")
health <- load_ssp("data/health.xlsx", "health")
gini <- load_ssp("data/gini.xlsx", "gini")
wap <- load_ssp("data/wap.xlsx", "wap")
tec <- load_ssp("data/tec.xlsx", "tec")

countries <-edu %>%  
  select(OID, GID_0) %>%
  distinct()
```

## Calculations

```{r output=FALSE}

#root mean square error (using median)
rmse_med <- function(x) {
  med = median(x, na.rm=TRUE)
  return(sum(sqrt((x-med)^2)))
}

#range function
rangesr <- function(x) {
  maxv = max(x, na.rm=TRUE)
  minv = min(x, na.rm=TRUE)
  return(diff(c(minv,maxv)))
}

out_all = countries
for(nm in list(edu,ma, gdp, health, gini, wap, tec)){
  rmsemed_nm = paste0(nm$variable[1],"_rmsemed")   
  medmed_nm = paste0(nm$variable[1],"_medmed")
  summed_nm = paste0(nm$variable[1],"_summed")
  sumrng_nm = paste0(nm$variable[1],"_sumrng")
  summax_nm = paste0(nm$variable[1],"_summax")
  summin_nm = paste0(nm$variable[1],"_summin")
  
  out_all <- nm %>%
  #edu %>%
    filter(unit=='rank') %>%  
    filter(!is.na(value)) %>%
    group_by(GID_0, ssp) %>%  
    #calc median rank across years, for each ssp, for each country
    summarise(medrank = median(value, na.rm=TRUE),
              maxrank = max(value, na.rm=TRUE),
              minrank = min(value, na.rm=TRUE),
              rangerank = rangesr(value),
              ) %>%  
    group_by(GID_0) %>%
    summarise(!!rmsemed_nm := rmse_med(medrank), #calc rmse (median) for ssp medians 
              !!medmed_nm := median(medrank),    #calc median of ssp medians 
              !!summed_nm := sum(medrank),       #calc sum of ssp medians
              !!sumrng_nm := sum(rangerank),     #calc sum of ssp ranges
              !!summax_nm := sum(maxrank),       #calc sum of ssp maxs
              !!summin_nm := sum(minrank),       #calc sum of ssp mins
              ) %>%   
    
    left_join(out_all, ., by='GID_0')
}


out_2100 = countries
for(nm in list(edu,ma, gdp, health, gini, wap, tec)){
  rmsemed_nm = paste0(nm$variable[1],"_rmsemed")  #calc rmse (median) for ssp medians  
  med_nm = paste0(nm$variable[1],"_med")
  rng_nm = paste0(nm$variable[1],"_rng")
  max_nm = paste0(nm$variable[1],"_max")
  min_nm = paste0(nm$variable[1],"_min")
  sum_nm = paste0(nm$variable[1],"_sum")
  
  out_2100 <- nm %>%
  #edu %>%
    filter(unit=='rank', year==2100)  %>%  
    filter(!is.na(value)) %>%
    group_by(GID_0) %>%
    summarise(!!rmsemed_nm := rmse_med(value),       ##calc rmse (median) for ssp ranks  
              !!med_nm := median(value, na.rm=TRUE), #calc median of ssp 2100 ranks 
              !!sum_nm := sum(value),                #calc sum of ssp 2100 ranks
              !!rng_nm := rangesr(value),            #calc range of ssp 2100 ranks
              !!max_nm := max(value, na.rm=TRUE),    #calc sum of ssp 2100 rank maxs
              !!min_nm := min(value, na.rm=TRUE),    #calc sum of ssp 2100 rank mins
              ) %>%   
    
    left_join(out_2100, ., by='GID_0')
}
```

## Plots

```{r}
#setup template https://stackoverflow.com/a/16727357

pall <- 
  list(
    geom_point(alpha=0.75),
    ggtitle("SSPs over all Years"),
    #scale_colour_gradient(low="darkblue",high="darkgrey"),
    scale_colour_viridis_c(),
    xlim(-5, NA),
    ylim(-5, NA),
    geom_abline(intercept = 0, slope = 1, 
                linewidth = 0.35,colour='red', linetype='dashed'),
    geom_label_repel(aes(label = GID_0),
                    alpha=0.5,
                    max.overlaps=15,
                    box.padding   = 0.35, 
                    point.padding = 0.5,
                    segment.color = 'grey50',
                    show.legend = FALSE,
                    color='black'),
    theme_classic()
  )

```


```{r}
#setup template https://stackoverflow.com/a/16727357

p2100 <- 
  list(
    geom_point(alpha=0.75),
    ggtitle("SSP Rankings in 2100"),
    #scale_colour_gradient(low="darkblue",high="darkgrey"),
    scale_colour_viridis_c(),
    xlim(-5, NA),
    ylim(-5, NA),
    geom_abline(intercept = 0, slope = 1, 
                linewidth = 0.35,colour='red', linetype='dashed'),
    geom_label_repel(aes(label = GID_0),
                    alpha=0.5,
                    max.overlaps=10,
                    max.time=1,
                    box.padding   = 0.35, 
                    point.padding = 0.5,
                    segment.color = 'grey50',
                    min.segment.length = (unit(0.01, "cm")),
                    show.legend = FALSE,
                    color='black',
                    size=4),
    theme_classic()
  )

```


### Education


#### All Years


```{r warning=FALSE}
out_all %>%
  drop_na() %>%
  ggplot(aes(x=edu_rmsemed, y=gdp_rmsemed, size=edu_summed, colour=gdp_summed)) +
  pall

ggsave("svg/edu_all.svg")
```

More grey is higher overall GDP ranking (e.g. USA is in bottom left)

#### 2100

```{r warning=FALSE}
out_2100 %>%
  drop_na() %>%
  ggplot(aes(x=edu_rng, y=gdp_rng, size=edu_med, colour=gdp_med)) +
  p2100 +
  ylab("GDP (range)") +
  xlab("Education (range)") +
  labs(size="Education (median)", colour="GDP (median)")

ggsave("svg/edu2100.svg")
```

More grey is higher overall GDP ranking (e.g. USA is in bottom left)

**Interpretation**

-   China has low GDP range (always ranked well in 2100), but relatively variable Educational rank

-   USA always ranked well on both indicators

-   Sri Lanka (LKA) has a relatively consistent GDP ranking (quite high), but highly variable education ranking

-   Mozambique, Burundi, Myanmar have variable GDP ranking (intermediate), but relatively consistent (poor) education ranking


### Health
```{r warning=FALSE}
out_all %>%
  drop_na() %>%
  ggplot(aes(x=health_rmsemed, y=gdp_rmsemed, size=health_summed, colour=gdp_summed)) +
  pall
```

```{r warning=FALSE}
out_2100 %>%
  drop_na() %>%
  ggplot(aes(x=health_rng, y=gdp_rng, size=health_med, colour=gdp_med)) +
  p2100 +
  ylab("GDP (range)") +
  xlab("Health (range)") +
  labs(size="Health (median)", colour="GDP (median)")

ggsave("svg/health2100.svg")
```
**Interpretation**

-   USA has low GDP range (always ranked well in 2100), but relatively variable health rank (compared to e.g India and so USA has higher median rank)

-   India always ranked well on both indicators (with low range)

-   Sri Lanka (LKA) and Belize (BLZ) are located similarly for both health and gdp range (high and low respectively), but BLZ has lower GDP and health median rankings

-   Mozambique, Burundi, Myanmar have variable GDP ranking (intermediate), but relatively consistent (quite low) health ranking

-  Bolivia (?BOL) has both extremely high range of health and GDP - are these correlated for BOL? (If so, why for this country and not others?)

### Market Access

```{r warning=FALSE}
out_all %>%
  drop_na() %>%
  ggplot(aes(x=ma_rmsemed, y=gdp_rmsemed, size=ma_summed, colour=gdp_summed)) +
  pall
```

```{r warning=FALSE}
out_2100 %>%
  drop_na() %>%
  ggplot(aes(x=ma_rng, y=gdp_rng, size=ma_med, colour=gdp_med)) +
  p2100 +
  ylab("GDP (range)") +
  xlab("Market Access (range)") +
  labs(size="MA (median)", colour="GDP (median)")

ggsave("svg/ma2100.svg")
```


**Interpretation**

-   USA has low GDP range (always ranked well in 2100), but relatively variable market access) rank (compared to e.g China nd India) but with relatively low median rank. What's going on here? Is there on particular SSP that USA ranks poorly on? (if so, why?)

-   China (and India) always ranked well on both indicators (with low range)

-   Germany (DEU) has very high market access range (but low GDP range) - this is because of SSP5 interpretation (we can see this in the pixel-level maps for example). Where are GNB and GMB? Do they stand out in the pixel-level maps too? 


-   Mozambique, Burundi, Myanmar have variable GDP ranking (intermediate), but relatively consistent (quite low) health ranking

-  In contrast to Health, Bolivia (?BOL) has low range for Market access (always low). Similarly highly variable GDP 


### Gini

```{r warning=FALSE}
out_all %>%
  drop_na() %>%
  ggplot(aes(x=gini_rmsemed, y=gdp_rmsemed, size=gini_summed, colour=gdp_summed)) +
  pall
```

```{r warning=FALSE}
out_2100 %>%
  drop_na() %>%
  ggplot(aes(x=gini_rng, y=gdp_rng, size=gini_med, colour=gdp_med)) +
  p2100 +
  ylab("GDP (range)") +
  xlab("Gini (range)") +
  labs(size="Gini (median)", colour="GDP (median)")

ggsave("svg/gini2100.svg")
```
This is a mess! 

Gini has the greatest range of values by country, >150 across the SSPs for MEX, PHL and GTM (all with quite low median and high GDP, so there must be one SSP in which they rank much higher than the rest) 

Interestingly this is where we see some of the greatest ranges for countries with low median ranks. Shows how while countries may look good on GDP for most SSPs, there are odd SSPs where the dsitribution of wealth becomes very unequal 


### Working Age Population

```{r warning=FALSE}
out_all %>%
  drop_na() %>%
  ggplot(aes(x=wap_rmsemed, y=gdp_rmsemed, size=wap_summed, colour=gdp_summed)) +
  pall
```

```{r warning=FALSE}
out_2100 %>%
  drop_na() %>%
  ggplot(aes(x=wap_rng, y=gdp_rng, size=wap_med, colour=gdp_med)) +
  p2100 +
  ylab("GDP (range)") +
  xlab("Working Age Population (range)") +
  labs(size="WAP (median)", colour="GDP (median)")

ggsave("svg/wap2100.svg")
```

WAP has some countries with very different position in 2100 vs over all years (e.g. USA, ESH, MMR). Why?

Otherwise, countries with generally higher GDP rank (and low range) such as FRA and DEU have higher WAP range. In contrast with the reverse for countres like MOZ and BDI 


### Energy

```{r warning=FALSE}
out_all %>%
  drop_na() %>%
  ggplot(aes(x=tec_rmsemed, y=gdp_rmsemed, size=tec_summed, colour=gdp_summed)) +
  pall
```

```{r warning=FALSE}
out_2100 %>%
  drop_na() %>%
  ggplot(aes(x=tec_rng, y=gdp_rng, size=tec_med, colour=gdp_med)) +
  p2100  +
  ylab("GDP (range)") +
  xlab("Energy (range)") +
  labs(size="Energy (median)", colour="GDP (median)")

ggsave("svg/tec2100.svg")
```

LUX, Israel, Slovenia, Estonia stand out with high ranges on energy use. Why?

India, China, USA highly ranked and low range on both. 

Note that LUX has such low median GDP - because it's so small (would rank near the top for per capita)










**Experimental**
Another way to plot these (if GDP is always the comparator) might be to show countries ranked on y-axis by gdp (med rank) then points on xaxis for the other variable value (for each ssp), then facet on x for variables (but this then does not show countries with variable vs non-variable gdp)

However, the below shows that with 200+ countries it's hard to get this looking good. It does show a string relationship between GDP and WAP (and to some degree energy). 

```{r}

#create data for 2100 to trial gdp ranked plot

ranks_2100 = edu
counter = 1
for(nm in list(edu,ma, health, gini, wap, tec)){
  
  if(counter == 1){ 
    ranks_2100 <- filter(nm, unit=='rank', year==2100)  
  } else {
    ranks_2100 <- bind_rows(ranks_2100, filter(nm, unit=='rank', year==2100))
  }
  counter = counter + 1
}

#create gdp rank data
gdp_2100 <-
  out_2100 %>%
  select(OID, GID_0, gdp_med)

#join ranks for our metrics to gdp 
gdp_ranks_2100 <- 
  left_join(gdp_2100, ranks_2100, by='GID_0',suffix=c("",".y")) %>%
  select(-ends_with(".y"))

```


```{r fig.height = 14}

#plot for EDU
gdp_ranks_2100 %>%
  drop_na() %>%
  arrange(gdp_med) %>%  #order countries by median SSP GDP 
  #filter(variable=='edu') %>%
  ggplot(aes(x=value, y=reorder(GID_0,gdp_med), colour=ssp)) +
  geom_point(alpha=0.75) +
  theme(axis.text.y = element_text(size=rel(0.65))) +
  facet_grid(.~variable) 

  
```  
