---
title: "SSPs Human Capitals - National"
author: "James Millington"
format: pdf
editor: visual
---

## Libraries

```{r}
library(tidyr)
library(dplyr)
library(readxl)
library(ggplot2)
library(ggrepel)
library(ggiraph)
```

## Load Data

```{r}
load_ssp <- function(filepath,varlab){
  #for each SSP data sheet
  for(n in 1:5){
    ssp <- paste0("ssp", n)
    sheet <- read_excel(filepath, sheet = ssp, na='<Null>')
    
    #pivot longer, add ssp id column
    lsheet <- sheet %>%
      pivot_longer(
        cols = !OID:unit,
        names_to='year',
        values_to='value'
      ) %>%
      mutate(ssp=n)
    
    #calculate ranks of values 
    lranks <- lsheet %>%
      group_by(year) %>%
      mutate(unit='rank',
             #value=min_rank(desc(value)))   #high rank number is low value
             value=min_rank(value))   #high rank number is high value
  
    #combine into single dataframe  
    ldat <- rbind(lsheet,lranks)
    if(n > 1){
      alldat <- rbind(alldat,ldat)
    } else { alldat <- ldat}
  }
  alldat <- mutate(alldat,variable=varlab)
  return(alldat)
}
```

```{r}
edu <- load_ssp("data/edu.xlsx", "edu")
ma <- load_ssp("data/ma.xlsx", "ma")
gdp <- load_ssp("data/gdp.xlsx", "gdp")
health <- load_ssp("data/health.xlsx", "health")
gini <- load_ssp("data/gini.xlsx", "gini")
wap <- load_ssp("data/wap.xlsx", "wap")
tec <- load_ssp("data/tec.xlsx", "tec")

countries <-edu %>%  
  select(OID, GID_0) %>%
  distinct()
```

## Calculations

```{r}

#root mean square error (using median)
rmse_med <- function(x) {
  med = median(x, na.rm=TRUE)
  return(sum(sqrt((x-med)^2)))
}

#range function
rangesr <- function(x) {
  max = max(x, na.rm=TRUE)
  min = min(x, na.rm=TRUE)
  return(diff(c(min,max)))
}

out_all = countries
for(nm in list(edu,ma, gdp, health, gini, wap, tec)){
  rmsemed_nm = paste0(nm$variable[1],"_rmsemed")   
  medmed_nm = paste0(nm$variable[1],"_medmed")
  summed_nm = paste0(nm$variable[1],"_summed")
  sumrng_nm = paste0(nm$variable[1],"_sumrng")
  summax_nm = paste0(nm$variable[1],"_summax")
  summin_nm = paste0(nm$variable[1],"_summed")
  
  out_all <- nm %>%
  #edu %>%
    filter(unit=='rank') %>%  
    group_by(GID_0, ssp) %>%  
    #calc median rank across years, for each ssp, for each country
    summarise(medrank = median(value, na.rm=TRUE),
              maxrank = max(value, na.rm=TRUE),
              minrank = min(value, na.rm=TRUE),
              rangerank = rangesr(value)
              ) %>%  
    group_by(GID_0) %>%
    summarise(!!rmsemed_nm := rmse_med(medrank), #calc rmse (median) for ssp medians 
              !!medmed_nm := median(medrank),    #calc median of ssp medians 
              !!summed_nm := sum(medrank),       #calc sum of ssp medians
              !!sumrng_nm := sum(rangerank),     #calc sum of ssp ranges
              !!summax_nm := sum(maxrank),       #calc sum of ssp maxs
              !!summin_nm := sum(minrank),       #calc sum of ssp mins
              ) %>%   
    
    left_join(out_all, ., by='GID_0')
}


out_2100 = countries
for(nm in list(edu,ma, gdp, health, gini, wap, tec)){
  rmsemed_nm = paste0(nm$variable[1],"_rmsemed")  #calc rmse (median) for ssp medians  
  med_nm = paste0(nm$variable[1],"_med")
  rng_nm = paste0(nm$variable[1],"_rng")
  max_nm = paste0(nm$variable[1],"_max")
  min_nm = paste0(nm$variable[1],"_min")
  sum_nm = paste0(nm$variable[1],"_sum")
  
  out_2100 <- nm %>%
  #edu %>%
    filter(unit=='rank', year==2100)  %>%  
    group_by(GID_0) %>%
    summarise(!!rmsemed_nm := rmse_med(value),       ##calc rmse (median) for ssp ranks  
              !!med_nm := median(value, na.rm=TRUE), #calc median of ssp 2100 ranks 
              !!sum_nm := sum(value),                #calc sum of ssp 2100 ranks
              !!rng_nm := rangesr(value),            #calc range of ssp 2100 ranks
              !!max_nm := max(value, na.rm=TRUE),    #calc sum of ssp 2100 rank maxs
              !!min_nm := min(value, na.rm=TRUE),    #calc sum of ssp 2100 rank mins
              ) %>%   
    
    left_join(out_2100, ., by='GID_0')
}
```

## Plots

### All years

#### Education

```{r}

p <- out_all %>%
  drop_na() %>%
  ggplot(aes(x=edu_rmsemed, y=gdp_rmsemed, size=edu_summed, colour=gdp_summed)) +
  geom_point_interactive(alpha=0.5) +
  #geom_point_interactive(aes(tooltip = GID_0, data_id=GID_0),alpha=0.5) +
  ggtitle("All years") +
  scale_colour_gradient(low="darkblue",high="darkgrey")+
  xlim(-5, NA) +
  ylim(-5, NA) +
  geom_abline(intercept = 0, slope = 1, 
              linewidth = 0.35,colour='red', linetype='dashed') +
  geom_label_repel(aes(label = GID_0),
                  alpha=0.5,
                  max.overlaps=15,
                  box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'grey50',
                  show.legend = FALSE,
                  color='black') +
  theme_classic()

p
#girafe(ggobj = p)
```

More grey is higher overall GDP ranking (e.g. USA is in bottom left)

### 2100

```{r}
#setup template https://stackoverflow.com/a/16727357

p2100 <- 
  list(
    geom_point(alpha=0.75),
    ggtitle("SSPs in 2100"),
    scale_colour_gradient(low="darkblue",high="darkgrey"),
    xlim(-5, NA),
    ylim(-5, NA),
    geom_abline(intercept = 0, slope = 1, 
                linewidth = 0.35,colour='red', linetype='dashed'),
    geom_label_repel(aes(label = GID_0),
                    alpha=0.5,
                    max.overlaps=15,
                    box.padding   = 0.35, 
                    point.padding = 0.5,
                    segment.color = 'grey50',
                    show.legend = FALSE,
                    color='black'),
    theme_classic()
  )

```
#### Education

```{r}
out_2100 %>%
  drop_na() %>%
  ggplot(aes(x=edu_rng, y=gdp_rng, size=edu_med, colour=gdp_med)) +
  p2100

```

More grey is higher overall GDP ranking (e.g. USA is in bottom left)

**Interpretation**

-   China has low GDP range (always ranked well in 2100), but relatively variable Educational rank

-   USA always ranked well on both indicators

-   Sri Lanka (LKA) has a relatively consistent GDP ranking (quite high), but highly variable education ranking

-   Mozambique, Burundi, Myanmar have variable GDP ranking (intermediate), but relatively consistent (poor) education ranking

Another way to plot these (if GDP is always the comparator) would be to show countries ranked on y-axis by gdp (med rank) then points on xaxis for the other variable value (for each ssp), then facet on x for variables (but this then does not show countries with variable vs non-variable gdp)

#### Health
```{r}
out_2100 %>%
  drop_na() %>%
  ggplot(aes(x=health_rng, y=gdp_rng, size=health_med, colour=gdp_med)) +
  p2100
```

```{r}
out_2100 %>%
  drop_na() %>%
  ggplot(aes(x=health_rng, y=gdp_rng, size=health_med, colour=gdp_med)) +
  p2100
```

#### Market Access

```{r}
out_2100 %>%
  drop_na() %>%
  ggplot(aes(x=ma_rng, y=gdp_rng, size=ma_med, colour=gdp_med)) +
  p2100

```

#### Gini

```{r}
out_2100 %>%
  drop_na() %>%
  ggplot(aes(x=gini_rng, y=gdp_rng, size=gini_med, colour=gdp_med)) +
  p2100

```

#### Working Age Population

```{r}
out_2100 %>%
  drop_na() %>%
  ggplot(aes(x=wap_rng, y=gdp_rng, size=wap_med, colour=gdp_med)) +
  p2100
```

#### Energy

```{r}
out_2100 %>%
  drop_na() %>%
  ggplot(aes(x=tec_rng, y=gdp_rng, size=tec_med, colour=gdp_med)) +
  p2100
```
